---
title: "Survival Analysis: Heart Failure Clinical Records"
author: "Aaron Pongsugree"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
required_packages <- c("survival", "survminer", "tidyverse",
                        "gtsummary", "knitr", "kableExtra", "ggplot2", "broom")
missing_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if (length(missing_packages) > 0) install.packages(missing_packages, dependencies = TRUE)

knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

---

# 1. Introduction

Heart failure is a serious cardiovascular condition affecting millions worldwide.
This analysis applies survival analysis techniques to the
**Heart Failure Clinical Records Dataset** (Chicco & Jurman, 2020) to identify
clinical predictors of mortality.

**Objectives:**

- Estimate survival functions using Kaplan-Meier (KM) curves
- Compare survival between clinically relevant subgroups via log-rank tests
- Identify independent predictors of death using a Cox Proportional Hazards model

---

# 2. Data Overview

The dataset contains **299 patients** with heart failure, recruited at the
Faisalabad Institute of Cardiology (Pakistan, 2015). The follow-up period
ranged from 4 to 285 days.

| Variable | Description | Type |
|---|---|---|
| `age` | Patient age (years) | Continuous |
| `anaemia` | Decrease in red blood cells | Binary |
| `creatinine_phosphokinase` | CPK enzyme level (mcg/L) | Continuous |
| `diabetes` | Presence of diabetes | Binary |
| `ejection_fraction` | % blood leaving heart per contraction | Continuous |
| `high_blood_pressure` | Presence of hypertension | Binary |
| `platelets` | Platelets in blood (kiloplatelets/mL) | Continuous |
| `serum_creatinine` | Serum creatinine level (mg/dL) | Continuous |
| `serum_sodium` | Serum sodium level (mEq/L) | Continuous |
| `sex` | Patient sex (1 = Male, 0 = Female) | Binary |
| `smoking` | Smoking status | Binary |
| `time` | Follow-up period (days) | Continuous |
| `DEATH_EVENT` | Death during follow-up (1 = Yes) | Binary (outcome) |

---

# 3. Setup & Data Loading

```{r libraries}
library(survival)
library(survminer)
library(tidyverse)
library(gtsummary)
library(knitr)
library(kableExtra)
library(ggplot2)
```

```{r load-data}
df <- read.csv("data/heart_failure_clinical_records_dataset.csv")

# Factor binary variables for readability
df <- df %>%
  mutate(
    anaemia          = factor(anaemia,          labels = c("No", "Yes")),
    diabetes         = factor(diabetes,         labels = c("No", "Yes")),
    high_blood_pressure = factor(high_blood_pressure, labels = c("No", "Yes")),
    sex              = factor(sex,              labels = c("Female", "Male")),
    smoking          = factor(smoking,          labels = c("No", "Yes")),
    DEATH_EVENT      = as.numeric(DEATH_EVENT),
    age_group        = cut(age, breaks = c(0, 60, 70, 100),
                           labels = c("< 60", "60–70", "> 70"))
  )


```

---

# 4. Descriptive Statistics

```{r summary-table}
df %>%
  select(-time, -DEATH_EVENT) %>%
  tbl_summary(
    by = NULL,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2
  ) %>%
  bold_labels() %>%
  as_kable_extra(caption = "Table 1. Patient Characteristics (N = 299)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r event-summary}
df %>%
  summarise(
    `Total Patients`          = n(),
    Deaths                    = sum(DEATH_EVENT),
    `Death Rate (%)`          = round(mean(DEATH_EVENT) * 100, 1),
    `Median Follow-up (days)` = median(time)
  ) %>%
  kable(caption = "Table 2. Event Summary") %>%
  kable_styling(bootstrap_options = "bordered", full_width = FALSE)
```

*Note: Median survival was not reached during the study period, as fewer than
50% of patients experienced the event (death rate = 32.1%).*

---

# 5. Kaplan-Meier Survival Estimates

## 5.1 Overall Survival

```{r km-overall}
surv_obj <- Surv(time = df$time, event = df$DEATH_EVENT)

km_overall <- survfit(surv_obj ~ 1, data = df)

ggsurvplot(
  km_overall,
  data          = df,
  conf.int      = TRUE,
  risk.table    = TRUE,
  xlab          = "Follow-up Time (Days)",
  ylab          = "Survival Probability",
  title         = "Figure 1. Overall Kaplan-Meier Survival Curve",
  ggtheme       = theme_minimal(),
  palette       = "#2E86AB",
  surv.median.line = "hv"
)
```

## 5.2 Survival by Ejection Fraction Group

```{r km-ef}
df <- df %>%
  mutate(ef_group = cut(ejection_fraction,
                        breaks = c(0, 30, 45, 100),
                        labels = c("Low (≤30%)", "Mid (31–45%)", "High (>45%)")))

km_ef <- survfit(surv_obj ~ ef_group, data = df)

ggsurvplot(
  km_ef,
  data       = df,
  conf.int   = FALSE,
  risk.table = TRUE,
  pval       = TRUE,
  xlab       = "Follow-up Time (Days)",
  ylab       = "Survival Probability",
  title      = "Figure 2. Survival by Ejection Fraction Group",
  legend.title  = "EF Group",
  ggtheme    = theme_minimal(),
  palette    = c("#E74C3C", "#F39C12", "#27AE60")
)
```

## 5.3 Survival by Serum Creatinine Group

```{r km-creatinine}
df <- df %>%
  mutate(creatinine_group = cut(serum_creatinine,
                                breaks = c(0, 1.2, 2, Inf),
                                labels = c("Normal (≤1.2)", "Elevated (1.2–2)", "High (>2)")))

km_creat <- survfit(surv_obj ~ creatinine_group, data = df)

ggsurvplot(
  km_creat,
  data       = df,
  conf.int   = FALSE,
  risk.table = TRUE,
  pval       = TRUE,
  xlab       = "Follow-up Time (Days)",
  ylab       = "Survival Probability",
  title      = "Figure 3. Survival by Serum Creatinine Group",
  legend.title  = "Creatinine",
  ggtheme    = theme_minimal(),
  palette    = c("#27AE60", "#F39C12", "#E74C3C")
)
```

## 5.4 Survival by Sex

```{r km-sex}
km_sex <- survfit(surv_obj ~ sex, data = df)

ggsurvplot(
  km_sex,
  data       = df,
  conf.int   = TRUE,
  risk.table = TRUE,
  pval       = TRUE,
  xlab       = "Follow-up Time (Days)",
  ylab       = "Survival Probability",
  title      = "Figure 4. Survival by Sex",
  legend.title  = "Sex",
  ggtheme    = theme_minimal(),
  palette    = c("#9B59B6", "#3498DB")
)
```

---

# 6. Log-Rank Tests

```{r logrank-tests}
comparisons <- list(
  "Ejection Fraction Group" = survdiff(surv_obj ~ ef_group,             data = df),
  "Serum Creatinine Group"  = survdiff(surv_obj ~ creatinine_group,     data = df),
  "Age Group"               = survdiff(surv_obj ~ age_group,            data = df),
  "Sex"                     = survdiff(surv_obj ~ sex,                  data = df),
  "Anaemia"                 = survdiff(surv_obj ~ anaemia,              data = df),
  "Diabetes"                = survdiff(surv_obj ~ diabetes,             data = df),
  "High Blood Pressure"     = survdiff(surv_obj ~ high_blood_pressure,  data = df),
  "Smoking"                 = survdiff(surv_obj ~ smoking,              data = df)
)

logrank_results <- tibble(
  Variable = names(comparisons),
  Chi_sq   = sapply(comparisons, function(x) round(x$chisq, 3)),
  df       = sapply(comparisons, function(x) length(x$n) - 1),
  p_value  = sapply(comparisons, function(x) round(1 - pchisq(x$chisq, length(x$n) - 1), 4))
)

logrank_sorted <- logrank_results %>%
  arrange(p_value) %>%
  mutate(Significant = ifelse(p_value < 0.05, "✓", ""))

logrank_sorted %>%
  kable(caption = "Table 3. Log-Rank Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(logrank_sorted$p_value < 0.05), bold = TRUE, color = "#27AE60")
```

---

# 7. Cox Proportional Hazards Model

## 7.1 Univariable Cox Models

```{r cox-univar}
predictors <- c("age", "anaemia", "creatinine_phosphokinase", "diabetes",
                "ejection_fraction", "high_blood_pressure", "platelets",
                "serum_creatinine", "serum_sodium", "sex", "smoking")

uni_results <- map_dfr(predictors, function(var) {
  formula <- as.formula(paste("surv_obj ~", var))
  model   <- coxph(formula, data = df)
  s       <- summary(model)$coefficients
  tibble(
    Variable = var,
    HR       = round(exp(s[1, "coef"]), 3),
    CI_lower = round(exp(s[1, "coef"] - 1.96 * s[1, "se(coef)"]), 3),
    CI_upper = round(exp(s[1, "coef"] + 1.96 * s[1, "se(coef)"]), 3),
    p_value  = round(s[1, "Pr(>|z|)"], 4)
  )
})

uni_sorted <- uni_results %>%
  arrange(p_value) %>%
  mutate(`95% CI` = paste0("(", CI_lower, " – ", CI_upper, ")")) %>%
  select(Variable, HR, `95% CI`, p_value)

uni_sorted %>%
  mutate(p_value = ifelse(p_value < 0.001, "<0.001", as.character(p_value))) %>%
  kable(caption = "Table 4. Univariable Cox Regression Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(uni_sorted$p_value < 0.05), bold = TRUE)
```

## 7.2 Multivariable Cox Model

Variables with p < 0.10 in univariable analysis are included.

```{r cox-multi}
cox_multi <- coxph(
  surv_obj ~ age + ejection_fraction + serum_creatinine + 
             serum_sodium + high_blood_pressure + anaemia,
  data = df
)
```

```{r cox-multi-table}
broom::tidy(cox_multi, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  mutate(`p.value` = ifelse(p.value < 0.001, "<0.001", as.character(p.value))) %>%
  mutate(term = recode(term,
    "high_blood_pressureYes" = "High Blood Pressure (Yes)",
    "anaemiaYes"             = "Anaemia (Yes)",
    "age"                    = "Age",
    "ejection_fraction"      = "Ejection Fraction",
    "serum_creatinine"       = "Serum Creatinine",
    "serum_sodium"           = "Serum Sodium"
  )) %>%
  mutate(p.value = ifelse(p.value < 0.001, "<0.001", as.character(p.value))) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(Variable = term, HR = estimate,
         `95% CI Lower` = conf.low,
         `95% CI Upper` = conf.high,
         `p-value` = p.value) %>%
  kable(caption = "Table. Multivariable Cox Model Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r cox-forest}
ggforest(cox_multi,
         data     = df,
         main     = "Figure 5. Forest Plot — Multivariable Cox Model",
         fontsize = 0.85)
```

## 7.3 Proportional Hazards Assumption Check

```{r ph-assumption, results='hide'}
ph_test <- cox.zph(cox_multi)
print(ph_test)
ggcoxzph(ph_test)
```

---

# 8. Key Findings

```{r findings-table}
findings <- tibble(
  Finding = c(
    "32.1% mortality rate over a median follow-up of 115 days; median survival not reached",
    "Low ejection fraction (≤30%) significantly worsens survival",
    "Elevated serum creatinine is the strongest independent predictor of mortality",
    "Older age independently increases hazard of death",
    "High blood pressure independently associated with increased mortality risk",
    "Serum sodium shows a marginal inverse association with hazard (p = 0.068)"
  ),
  `Statistical Support` = c(
    "96 deaths among 299 patients; KM overall curve",
    "Log-rank p < 0.0001; HR = 0.954 per unit increase in multivariable Cox",
    "Log-rank p < 0.0001; HR = 1.356, p < 0.001 in multivariable Cox",
    "HR = 1.045 per year, p < 0.001 in multivariable Cox",
    "HR = 1.606, p = 0.026 in multivariable Cox",
    "HR = 0.958, p = 0.068; CI crosses 1.0"
  )
)

findings %>%
  kable(caption = "Table 5. Summary of Key Findings") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 9. Conclusion

This survival analysis identified **age**, **ejection fraction**, **serum creatinine**,
and **high blood pressure** as significant independent predictors of mortality in heart
failure patients (all p < 0.05), with serum sodium showing a borderline association
(p = 0.068). Serum creatinine carried the largest effect size (HR = 1.36), consistent
with Chicco & Jurman (2020). The global Schoenfeld test confirmed the proportional
hazards assumption was met (p = 0.207), though ejection fraction showed a marginal
individual violation (p = 0.033) warranting cautious interpretation. Overall model
discrimination was adequate (concordance = 0.73), supporting the clinical utility
of these variables in predicting heart failure mortality.

---

# 10. References

- Chicco, D., & Jurman, G. (2020). Machine learning can predict survival of
  patients with heart failure from serum creatinine and ejection fraction alone.
  *BMC Medical Informatics and Decision Making, 20*(16).
- Kaggle Dataset: Heart Failure Clinical Records —
  https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data


---
